Statically Typed String Sanitation Inside a Python (Technical Report)

### 2014-12-17 sanitation-psp14/tr/strings_chapter.tex
work.

\newcommand{\tcheck}[4]{{\sf check}(#1; #2; #3; #4)}
\renewcommand{\tstr}[1]{{{\sf str}[#1]}}
\newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...

@@

\item \textbf{SS-E-Replace}
Suppose that $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ 
The only rule that applies is S-T-Replace, so
$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
\begin{align*}
\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
\end{align*}

Therefore, it suffices to show that $\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$
or, equivalently, that $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
This follows from the fact that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
The only rule that applies is S-T-Replace, so
$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
\begin{align*}
\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
\emptyset &\vdash e_2 : \stringin{r_2}
\end{align*}
By induction, $\emptyset \vdash e_2' : \stringin{r_2}$.
Therefore, $\emptyset \vdash \rreplace{r}{\rstr{s_1}}{e_2'} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.

% bookmark-nrf
\item

\item \textbf{Cases SS-E-Replace-Left, SS-E-Replace-Right, SS-E-Check-StepLeft, SS-E-SafeCoerce-Step, SS-E-Check-StepRight}. \todo{todo}

@@

\begin{thm}[Semantic Correspondence]
If $\iota \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
\end{thm}
\begin{proof}
By induction on the structure of $\iota$.

\begin{itemize}[label=$ $,itemsep=1ex]
\item \todo{...}
\end{itemize}
\end{proof}

@@

We can also define canonical forms for regular expressions and strings in the usual way:

@@

\begin{thm}[Translation Correctness]\label{thm:trcorrect}
  If $\tctx \vdash e : \sigma$ then 
  there exists an $\iota$ such that $\trden{e} = \iota$
  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
  Furthermore, if $e \sreduces v$ then
  $\iota \treduces \dot{v}$ such that
  $\trden{v} = \dot{v}$.
\end{thm}
\begin{proof}
We present a proof by induction on the structure of $e$ \todo{why not on the typing derivation?}.
We write $e \leadsto \iota$ as shorthand for the final property.

@@

\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. This case is an obvious appeal to induction.

@@

@@

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
\end{mathpar}

@@

\inferrule[S-E-Replace]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}

>>>>

\newcommand{\tcheck}[4]{{\sf check}(#1; #2; #3; #4)}
\renewcommand{\tstr}[1]{{{\sf str}[#1]}}
\newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
\newcommand{\treplace}[3]{{\sf replace}(#1;#2;#3)}
\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...

@@

%\item \textbf{SS-E-Replace}
%Suppose that $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ 
%The only rule that applies is S-T-Replace, so
%$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
%\begin{align*}
%\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
%\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
%\end{align*}
%
%Therefore, it suffices to show that $\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$
%or, equivalently, that $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
%This follows from the fact that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
%%The only rule that applies is S-T-Replace, so
%%$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
%%\begin{align*}
%%\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
%%\emptyset &\vdash e_2 : \stringin{r_2}
%%\end{align*}
%%By induction, $\emptyset \vdash e_2' : \stringin{r_2}$.
%%Therefore, $\emptyset \vdash \rreplace{r}{\rstr{s_1}}{e_2'} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
%
%\item \textbf{SS-E-Check-StepLeft}.
%Suppose that $e = \rcheck{r}{e_0}{x.e_1}{e_2} \mapsto \rcheck{r}{e_0'}{x.e_1}{e_2}$.
%The only applicable typing rule is S-T-Check. 
%Therefore, $\emptyset \vdash e : \sigma$ where:
%\begin{align*}
%\emptyset &\vdash e_0 : \stringin{r} \\
%\emptyset, x:\stringin{r} &\vdash e_1 : \sigma \\
%\emptyset &\vdash e_2 : \sigma 
%\end{align*}
%By induction, $\emptyset \vdash e_0' : \stringin{r}$.
%Therefore, the result follows by P-T-Check.
%
%\item \textbf{SS-E-Check-Ok}.
%Suppose that $e = \rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \replace{\rstr{s}}{x}{e_1}$
%where $s \in \lang{r}$. The only applicable typing rule is S-T-Check. 
%Therefore, $\emptyset \vdash e : \sigma$ where:
%\begin{align*}
%\emptyset &\vdash e_0 : \stringin{r} \\
%\emptyset, x:\stringin{r} &\vdash e_1 : \sigma \\
%\emptyset &\vdash e_2 : \sigma 
%\end{align*}
%By substitution, $\emptyset \vdash \replace{\rstr{s}}{x}{e_1} : \sigma$.
%
%\item \textbf{SS-E-Check-NotOk}.
%Suppose that $e = \rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$
%where $s \not \in \lang{r}$. The only applicable typing rule is S-T-Check. 
%Therefore, $\emptyset \vdash e : \sigma$ where $\emptyset \vdash e_2 : \sigma$.

@@

\begin{thm}[Semantic Correspondence]
If $\iota \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
\end{thm}
\begin{proof}
By induction on the structure of $\iota$.
The proof is similar to the proof for $\lambdas$.
\todo{if there's time add the proof...}

%\begin{itemize}[label=$ $,itemsep=1ex]
%\item \todo{...}
%\end{itemize}
\end{proof}

@@

% We can also define canonical forms for regular expressions and strings in the usual way:

@@

\begin{thm}[Translation Correctness]\label{thm:trcorrect}
  If $\sctx \vdash e : \sigma$ then 
  there exists an $\iota$ such that $\trden{e} = \iota$
  and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.
  Furthermore, $e \sreduces v$ and
  $\iota \treduces \dot{v}$ such that
  $\trden{v} = \dot{v}$.
\end{thm}
\begin{proof}
We present a proof by induction on the structure of $e$.
We write $e \leadsto \iota$ as shorthand for the final property.

@@

\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. \todo{hand wave lots and lots of symbol pushing.} By induction.

@@

\item \textbf{Case $e = \rreplace{r}{e_1}{e_2}$}.
There is only one applicable typing rule, so suppose
$\sctx \vdash \rreplace{r}{e_1}{e_2} : \stringin{\lreplace{r}{e_1}{e_2}}$.
Let $\psi = \trden{\sctx}$.
Note that $\trden{\rreplace{r}{e_1}{e_2}} = \treplace{\rx{r}}{\iota_1}{\iota_2}$
when by induction $\trden{e_1} = \iota_1$ and $\trden{e_2} = \iota_2$
such that $\psi \vdash \iota_1$ and $\psi \vdash \iota_2$.
It follows by P-T-Replace that $\psi \vdash \treplace{\rx{r}}{\iota_1}{\iota_2} : \str$.
Finally, note that $\trden{\stringin{\lreplace{r}{e_1}{e_2}}} = \str$.

For evaluation correspondence, note that $\trden{ \rstr{\lreplace{r}{s_1}{s_2}} } = \rstr{ \lreplace{r}{s_1}{s_2} }$ and
so it suffices to show that $\treplace{\rx{r}}{\iota_1}{\iota_2} \sreduces \rstr{r}{s_1}{s_2}$.
Note that $\lreplace{r}{e_1}{e_2} \sreduces \rstr{\lreplace{r}{s_1}{s_2}}$ 
where $e_1 \sreduces \rstr{s_1}$, $e_2 \sreduces \rstr{s_2}$, $r \sreduces r$.
By induction, $\iota_1 \sreduces \rstr{s_1}, \iota_2 \sreduces \rstr{s_2}$, and $\rx{r} \sreduces \rx{r}$.
So by S-E-Replace, the sufficient condition holds.

\item \textbf{Case $e = \rcoerce{r}{e'}$}.
The only applicable tpying rule is S-T-SafeCoerce, so suppose
$\sctx \vdash \rcoerce{r}{e'} : \stringin{r}$ where
$\sctx \vdash e' : \stringin{r'}$ and $\lang{r'} \subseteq \lang{r}$.
By induction, $e' \leadsto \iota$ for some $\iota$.
Therefore, $\trden{\coerce{r}{e'}} = \iota$ by Tr-SafeCoerce.

For evaluation correspondence, note that $e \sreduces v$ where
$e' \sreduces v$. The result follows by induction because $e' \leadsto \iota$.

\item \textbf{Case $e = \rcheck{r}{e_1}{x.e_2}{e_3}$}.
The applicable typing rule is S-T-Check,
so $\psi \vdash e : \sigma$ where 
$\psi \vdash e_1 : \stringin{r}$,
$\psi, x : \stringin{r} \vdash e_2 : \sigma$,
and $\psi \vdash e_3 : \sigma$.
By induction and a corresponding substitution principle there exists $\iota_1, \iota_2, \iota_3$ such that
$e_1 \leadsto \iota_1$, 
$e_2 \leadsto \iota_2$ in context $\psi, s : \stringin{r}$, and
$e_3 \leadsto \iota_3$.
Choose $\iota = \tcheck{\rx{r}}{\iota_1}{\lambda x . \iota_2}{\iota_3}$.
\todo{hand wave} The result follows by induction.

@@

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
\end{mathpar}

@@

\inferrule[S-E-Replace]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} } 
{\sisubst{r}{e_1}{e_2} \sreduces \sistr{ \lsubst{r}{s_1}{s_2}  }}
